---
title: "Assignment 1 - Language Development in ASD - part 2"
author: "Riccardo Fusaroli"
date: "July 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Language development in Autism Spectrum Disorder (ASD)

Background: Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail: i) relying on actual naturalistic language production, ii) over extended periods of time. We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

We then want to test the language trajectory of child and parent over time.

This Markdown is structured in the following way:

1. The exercises: read them carefully. Under each exercise you will have to write your answers, once you have written and run the code. This is the part that you have to directly send to the teachers.
2. An (optional) guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results under the exercise part.
3. In exercise 4 you will be asked to create the best possible model of language development in TD and ASD children, picking and choosing whatever additional variables you want from the dataset. Next time, the models produced by the different groups will compete against each other to see who can produce the best model, so choose carefully!

You will have to have a github repository for the code and send the answers to Celine and Riccardo without code (but a link to your github repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

N.B. The following lines are a summary of the questions to be answered, the step-by-step instructions and tips are below.

## Exercise 1) Preliminary Data Exploration

Describe the participant samples in the dataset (e.g. by diagnosis, age, etc.). Do you think the two groups are well balanced? If not, what do you think was the reason?

The only parameter where the two groups are not well-matched is on age.
The age differs significantly between the two groups. Typically developing children in the sample were on average 12.6296 months (SE = 1.0251) younger than children with autism (t = -12.32, p < 0.05).
The balance of gender does not differ significantly across the two groups (z = 0.2940, p < 0.05)
The MLU at first visit does not differ significantly between the two groups at the first visit (t = 0.043, p > 0.05)
The Verbal IQ of the two groups does not differ significantly between the two groups (t = 1.79, p > 0.05)
The nonVerbalIQ of the two groups did not differ significantly between the two groups (t = -0.76, p > 0.05)

The two groups only differ significantly on their language abilities. On all other aspects (verbalIQ, nonverbalIQ, MLU at first visit and gender) there was no significant differences between the two groups.
The reason why the data had to be non-matched for age is that language development for ASD is usaually slower and therefore matching age and language fluency would have been impossible.

### Exercise 2) Children learning language: the effects of time and ASD
Describe linguistic development in TD and ASD children in terms of Mean Length of Utterance (MLU)?

Linguistic development of children MLU is affected significantly by time but not by diagnosis.
We produced a linear mixed effects analysis of the relationship between child MLU, time and diagnosis. As fixed effects we entered visit(time) and diagnosis. We also entered random slopes for each subject and random intercepts for visits as random effects.
MLU was not significantly affected by diagnosis (beta = 0.29, SE = 0.15, t = 1.882, p > 0.05).
However, MLU was significantly affected by visit (beta = 0.23, SE = 0.025, t = 9.437, p < 0.05), meaning that for each new visit MLU increases by 0.23.

Our model was not significant compared to a null-model including only the random effects and visit (chisq(1,7) = 2.0177, p > 0.05). THe full model accounted for 81 % of variance whereoff the fixed effects accounted for 22 % of variance (R^2M = 0.2153, R^2C = 0.8063). Thus adding diagnosis as predictor did not add predictive value to the model
The residuals were normally distributed.


We also made a test on an interaction model, where we checked the interaction between visit and diagnosis. This model was significantly better than the model without an interaction effect (chisq(1, 8) = 34.962, p < 0.05).
In the model visit was significant (beta = 0.1044, SE = 0.027, t = 3.686, p < 0.05). Diagnosis was still not significant (beta = -0.217, SE = 0.17, t = -1.260, p > 0.05)
The interaction however was significant (beta = 0.253, SE = 0.038, t = 6.715, p < 0.05). The model with interaction accounted for 82 % of variance in the data, whereoff the fixed effects accounted for 35 % (R^2C = 0.818, R^2M = 0.351)


### Exercise 3) Child directed speech as a moving target
Describe how parental use of language changes over time in terms of MLU. What do you think is going on?

MLU of the children's parents is affected significantly by time and by diagnosis.

We produced a linear mixed effects analysis of the relationship between parent MLU, time and diagnosis. As fixed effects we entered visit(time) and diagnosis. We also entered random slopes for each subject and random intercepts for visits as random effects.
MLU of the parents was significantly affected by diagnosis (beta = 0.502, SE = 0.12, t = 6.542, p < 0.05), meaning that parents with a typically developing child had a MLU that was 0.502 longer than parents of children with ASD.
MLU of the parents was significantly affected by visit (beta = 0.12, SE = 0.02, t = 6.542, p < 0.05), meaning that for each new visit MLU increases by 0.12.
The model explained 68 percent of variance and the fixed effects explained 23 % of variance (R^2C = 0.6, R^2M = 0.23).

The residuals was normally distributed.

THe model was significantly better than a nullmodel, only including the random effects (chisq(2, 7) = 49.95, p < 0.05)

What is going on is that parents are increasing their MLU by time and that parents with children with ASD use a smaller MLU as their children normally use a smaller MLU as well.

### Exercise 4) Looking into "individual differences" (demographic, clinical or cognitive profiles)
The dataset contains some additional variables characterizing the kids’ cognitive and clinical profile: ADOS (autism severity), MSEL EL (Expressive Language, that is, verbal IQ, or linguistic skills at first visit as assessed by a psychologist using Mullen Scales of Early Learning), MSEL VR (Visual Reception, used as a proxy for non verbal IQ at first visit), Age, Gender, Ethnicity. Would it make sense to add any of them to your model of linguistic trajectories? Create the best possible model (the one that best explain the data, with MLU as outcome). Next time your model will be tested on new participants, and we will proclaim a winner. Describe your strategy to select the best models (how did you choose the variables to include?) and send the code to Riccardo and Celine.


A model including an interaction between ADOS and visit, as well as the variable verbalIQ could explain 82 % of the variance in the data whereoff the fixed effects explained 65 % of the variance (R^2C = 0.82, R^2M = 0.65).
MLU of the child was significantly affected by visit (beta = 0.37, SE = 0.02, t = 15.856, p < 0.05), which means that MLU increased by 0.37 per visit.
MLU of the child was significantly affected by ADOS (beta = 0.03, SE = 0.01, t = 3.607, p < 0.05), which means that children with one point higher ADOS had an MLU that was higher by 0.03 (this is probably due to a slight difference i the means and the interaction will still make the MLU of children with autism lower.)
MLU of the child was significantly affected by verbalIQ (beta = 0.07, sE = 0.01, t = 8.918, p < 0.05) meaning that the childs MLU was 0.07 higher when IQ increased by 1.
MLU of the child was significantly affected by the interaction between visit and ADOS (beta = -0.02, SE = 0.00, t = -8.549, p < 0.05).



### [OPTIONAL] Exercise 5) Comment on how the three linguistic variables measure linguistic performance (the so-called "construct validity" of the measures). Do they express the same variance?

The three linguistic measures are MLU, unique words and words said in totatl.
The developments of these 3 is probably correlated. However, MLU tells more about the complexity of a sentence, unique words are measuring which new words would have been learned and words in total how much the child speaks during a visit.
BUt they probably all express part of the same variance.

### Structure of the code chunks

Basic stuff:
- Loading the libraries
- Setting the directory and loading the data
- Look at the data (which variables are there? Are they in the right format?) and describe the participants (by diagnosis)

We will try to answer three questions:

- Do children with ASD develop language differently from non-ASD children?
- Do parents speak differently to children with ASD than to non-ASD ones?
- Which variables should we use to best explain the child linguistic performance?
  
### Loading the relevant libraries

Load necessary libraries : what will you need?

- e.g. something to plot with
- e.g. mixed effects models

```{r Load Libraries}

library(lmerTest)
library(MuMIn)
library(ggplot2)

```

### Define your working directory and load the data

- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data}

#getwd()
#locpath=
#setwd(locpath)
#Data =

setwd("C:/Users/hille/OneDrive/Cognitive Science at Aarhus University/2017 - Experimental Methods 3/assignments/Assignment-1")

#load data
data = read.csv("dataAssignment1.csv")

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Visit, Number of words used, Number of unique words used, length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r}
#Check difference in age between groups
#First violent plot
ggplot(data[data$VISIT == "1", ], aes(x = Diagnosis, y = Age))+
  geom_violin()+
  labs(title = "Difference in age between ASD and TD children", x = "Group", y = "Age")

summary(lm(Age ~ Diagnosis, data[data$VISIT == "1", ]))
#Significantly different



#Gender: Binomial (not continousess) outcome
summary(glm(Gender ~ Diagnosis, family = binomial, data[data$VISIT == "1",]))


#MLU at first visit
ggplot(data[data$VISIT == "1",], aes(Diagnosis, CHI_MLU))+
  geom_violin()+
  labs(title = "MLU at first visit", x = "Group", y = "MLU at first visit")

summary(lm(CHI_MLU ~ Diagnosis, data[data$VISIT == "1", ]))


#V_IQ
ggplot(data[data$VISIT == "1",], aes(Diagnosis, verbalIQ))+
  geom_violin()+
  labs(title = "Verbal IQ in the two groups", x = "group", y = "Verbal IQ")

summary(lm(verbalIQ ~ Diagnosis, data[data$VISIT == "1",]))


#NOnverbalIQ

ggplot(data[data$VISIT == "1",], aes(Diagnosis, nonVerbalIQ))+
  geom_violin()+
  labs(title = "Non verbal IQ in the two groups", x = "group", y = "Non verbal IQ")


summary(lm(nonVerbalIQ ~ Diagnosis, data[data$VISIT == "1", ]))


#Only difference is in Age

```

The only parameter where the two groups are not well-matched is on age.

The age differs significantly between the two groups. Typically developing children in the sample were on average 12.6296 months (SE = 1.0251) younger than children with autism (t = -12.32, p < 0.05).

The balance of gender does not differ significantly across the two groups (z = 0.2940, p < 0.05)

The MLU at first visit does not differ significantly between the two groups at the first visit (t = 0.043, p > 0.05)

The Verbal IQ of the two groups does not differ significantly between the two groups (t = 1.79, p > 0.05)

The nonVerbalIQ of the two groups did not differ significantly between the two groups (t = -0.76, p > 0.05)



## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r}
#Let us make a plot!
ggplot(data, aes(VISIT, CHI_MLU, colour = Diagnosis))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(title = "Relationship between time,  child MLU and diagnosis", x = "Visit", y = "Child MLU")+
  geom_jitter(width = 0.08)


#NUll model with only visit
nullModel = lmer(CHI_MLU ~ 1 + VISIT + (1 + VISIT|SUBJ), data)
nullModel
summary(nullModel)

#Model
model1 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1 + VISIT|SUBJ), data)


#A model with an interaction
model1interaction = lmer(CHI_MLU ~ 1 + VISIT*Diagnosis + (1 + VISIT|SUBJ), data)



```

How would you evaluate whether the model is a good model?
We would test it using an anova and find the variance explained by each of the models.

```{r}
#Compare model with null model using ANOVA
anova(nullModel, model1)


#ANOVA
anova(nullModel, model1interaction)


#How much variance is explained by each of the models?
#Find R^2
r.squaredGLMM(model1)

r.squaredGLMM(model1interaction)

#Testing the assumptions of normality of residuals
library(DHARMa)
simulationOutput = simulateResiduals(fittedModel = model1interaction, n = 250)

#Plot
plotSimulatedResiduals(simulationOutput = simulationOutput)

#Make a formal goodness-of-fit test
testUniformity(simulationOutput = simulationOutput)

# Non-significant


#Conceptual: Do we need these predictors in the models?


```

(we were told to skip this part as we did not cover it in class)
Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better

```{r}

```

Exciting right?
Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results

[REPORT THE RESULTS]
Linguistic development of children MLU is affected significantly by time but not by diagnosis.
We produced a linermixed effects analysis of the relationship between child MLU, time and diagnosis. As fixed effects we entered visit(time) and diagnosis. We also entered random slopes for each subject and random intercepts for visits as random effects.
MLU was not significantly affected by diagnosis (beta = 0.29, SE = 0.15, t = 1.882, p > 0.05).
However, MLU was significantly affected by visit (beta = 0.23, SE = 0.025, t = 9.437, p < 0.05), meaning that for each new visit MLU increases by 0.23.

Our model was not significant compared to a null-model including only the random effects and visit (chisq(1,7) = 2.0177, p > 0.05). The full model accounted for 81 % of variance whereoff the fixed effects accounted for 22 % of variance (R^2M = 0.2153, R^2C = 0.8063). Thus adding diagnosis as predictor did not add predictive value to the model
The residuals were normally distributed.


We also made a test on an interaction model, where we checked the interaction between visit and diagnosis. This model was significantly better than the model without an interaction effect (chisq(1, 8) = 34.962, p < 0.05).
In the model visit was significant (beta = 0.1044, SE = 0.027, t = 3.686, p < 0.05). Diagnosis was still not significant (beta = -0.217, SE = 0.17, t = -1.260, p > 0.05)
The interaction however was significant (beta = 0.253, SE = 0.038, t = 6.715, p < 0.05). The model with interaction accounted for 82 % of variance in the data, whereoff the fixed effects accounted for 35 % (R^2C = 0.818, R^2M = 0.351)

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r}
#Let us make a plot!
ggplot(data, aes(VISIT, MOT_MLU, colour = Diagnosis))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(title = "Relationship between time,  parent MLU and diagnosis", x = "Visit", y = "Parent MLU")+
  geom_jitter(width = 0.08)


#NUll model
nullModel2 = lmer(MOT_MLU ~ 1 + (1 + VISIT|SUBJ), data)
nullModel
summary(nullModel)

#Model
model2 = lmer(MOT_MLU ~ 1 + VISIT + Diagnosis + (1 + VISIT|SUBJ), data)

summary(model2)

#Interaction model
model2interaction = lmer(MOT_MLU ~ 1 + VISIT*Diagnosis + (1 + VISIT|SUBJ), data)

#Make an ANOVA comparing nullModel and model2
anova(nullModel2, model2)

#Make an ANOVA comparing model 2 and model2interaction
anova(model2, model2interaction)


#Testing r-squared
r.squaredGLMM(model2)

#Testing the assumptions of normality of residuals
simulationOutput2 = simulateResiduals(fittedModel = model2, n = 250)
plotSimulatedResiduals(simulationOutput = simulationOutput2)
testUniformity(simulationOutput = simulationOutput2)





```

MLU of the children's parents is affected significantly by time and by diagnosis.

We produced a linear mixed effects analysis of the relationship between parent MLU, time and diagnosis. As fixed effects we entered visit(time) and diagnosis. We also entered random slopes for each subject and random intercepts for visits as random effects.
MLU of the parents was significantly affected by diagnosis (beta = 0.502, SE = 0.12, t = 6.542, p < 0.05), meaning that parents with a typically developing child had a MLU that was 0.502 longer than parents of children with ASD.
MLU of the parents was significantly affected by visit (beta = 0.12, SE = 0.02, t = 6.542, p < 0.05), meaning that for each new visit MLU increases by 0.12.
The model explained 68 percent of variance and the fixed effects explained 23 % of variance (R^2C = 0.6, R^2M = 0.23).

The residuals was normally distributed.

THe model was significantly better than a nullmodel, only including the random effects (chisq(2, 7) = 49.95, p < 0.05)


### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Celine.

If we add enough variables we can explain a lot of variance. It does not make sense to make this as a complete exploratory analysis, therefore we will look upon which predictors might be predictive and make a model:

We already know that diagnosis in itself was not significant, but that an interaction between diagnosis and visit. ADOS(severity of autistic symptoms) would probably be more predictive than the diagnosis instead and therefore we want to swap ADOS and diagnosis. We want to keep the interaction, as there is probably an interaction between sevirity of autistic symptoms and time on development of languge. ADOS might also explain a great deal of variance. And would have been correlated with diagnosis, thus it does not make sense to keep both!

VerbalIQ would also very likely be predictive of ability to learn language over time.
NonVerbalIQ should not be predictive of language abilities.
It does not make sense to include any of the data on the mother or how the child performed in a clinical trial, because then the model predicting the future, would require collecting future data and you might also as well look directly at the MLU in the first place.

We will try it out adding variables one at a time testing significance.
Also start out with a null-model and so forth to see if adding each predictor makes the model significantly better by an ANOVA. But our goal is the model above.

```{r}

model3 = lmer(CHI_MLU ~ 1 + VISIT + (1 + VISIT|SUBJ), data)
model4 = lmer(CHI_MLU ~ 1 + VISIT + ADOS + (1 + VISIT|SUBJ), data)
model5 = lmer(CHI_MLU ~ 1 + VISIT*ADOS + (1 + VISIT|SUBJ), data)

#The model we wanted to reach
model6 = lmer(CHI_MLU ~ 1 + VISIT*ADOS + verbalIQ + (1 + VISIT|SUBJ), data)

#What do we do for an ANOVA?
#Check if every predictor makes the model more significant?
#Yeah, we kinda need to do that.

anova(model3, model4)
#Significant

anova(model4, model5)
#Significant

anova(model5, model6)


#Testing r-squared
r.squaredGLMM(model6)

#I am in doubt if I should report all of this?

```

REPORT OF THE RESULTS


A model including an interaction between ADOS and visit, as well as the variable verbalIQ could explain 82 % of the variance in the data whereoff the fixed effects explained 65 % of the variance (R^2C = 0.82, R^2M = 0.65).
MLU of the child was significantly affected by visit (beta = 0.37, SE = 0.02, t = 15.856, p < 0.05), which means that MLU increased by 0.37 per visit.
MLU of the child was significantly affected by ADOS (beta = 0.03, SE = 0.01, t = 3.607, p < 0.05), which means that children with one point higher ADOS had an MLU that was higher by 0.03 (this is probably due to a slight difference i the means and the interaction will still make the MLU of children with autism lower.)
MLU of the child was significantly affected by verbalIQ (beta = 0.07, sE = 0.01, t = 8.918, p < 0.05) meaning that the childs MLU was 0.07 higher when IQ increased by 1.
MLU of the child was significantly affected by the interaction between visit and ADOS (beta = -0.02, SE = 0.00, t = -8.549, p < 0.05).


